Why Donors Donate: Disentangling Organizational and Structural Heuristics for International Philanthropy

### 2025-01-18 manuscript/manuscript.qmd
Add Suparna edits

---
title: "Why Donors Donate: Disentangling Organizational and Structural Heuristics for International Philanthropy"
short-title: "Why Donors Donate"
date: "2025-01-11"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  url: http://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
  orcid: 0000-0003-3506-7590
- name: Marc Dotson
  affiliation: Utah State University
  url: https://occasionaldivergences.com/
  email: marc.dotson@usu.edu
  orcid: 0000-0001-7846-3660
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  Non-governmental organizations (NGOs) are a core component of global governance. Over the last two decades, however, many host governments have cracked down on NGOs using a series of laws that adversely affect the availability and use of traditional NGO funding. In response to this dramatic shift in the funding landscape, international NGOs have increasingly turned to individual donors to offset the loss funding. Prior research on NGO fundraising has examined how financial transparency and organizational accountability influence individual donor behavior. How do these elements of private donor decision-making interact with structural-level factors such as worsening host country civic environments when deciding to donate? Using a conjoint experiment with likely donors in the US, we find that transparency and accountability can protect against the dampening effect of host government crackdown and criticism of NGOs. Our results have important implications for how NGOs can adapt to worsening civic environments. 
additional-info: |
  **Data Availability:** All data and replication code are publicly available at <https://github.com/andrewheiss/silent-skywalk>
keywords:
  - philanthropy
  - conjoint experiments
  - donor heuristics
  - repression
  - NGOs
  - civil society
  - nonprofits
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/silent-skywalk>"
bibliography: references.bib
csl: csl/apa.csl
link-citations: true
word-count: "9,300"
---

@@

Non-governmental organizations (NGOs) face stricter regulations from their host governments [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020]. Repressive governments have cracked down on these organizations using both violence, as well as administrative crackdown, which create barriers to funding and advocacy for a variety of NGOs [@DupuyRonPrakash:2016; @Chaudhry:2022; @Heiss:2017]. Policymakers have referred to this phenomenon as the "closing of civic space" or "shrinking civic space" [@CarothersBrechenmacher:2014; @Rutzen:2015b]. Data from the CIVICUS Civil Society Monitor shows that civil society groups face serious restrictions in more than 60% of countries---117 in total [@CIVICUSMonitor]. These attacks---most of which have been implemented through anti-NGO laws---have adversely affected the ability of both domestic and international NGOs to operate and raise funds. Official aid funds have decreased substantially in countries that have imposed restrictive anti-NGO legislation [@ChaudhryHeiss:2018; @DupuyPrakash:2018] and official donors channel fewer funds to contentious issues such as human rights, elections, and advocacy, and instead choose to fund "tame" causes bush as health, education, and development, among others [@ChaudhryHeiss:2018; @Bush:2015; @Herrold:2020; @RightSpringmanWibbels:2022]. In response to this dramatic shift in the NGO funding landscape, international NGOs (INGOs) have increasingly turned to individual donors abroad to offset the loss of funding from government and foundation sources [@BanksHulmeEdwards:2015].

The amount of private foreign aid is not insignificant---from 2000--2010, private giving by U.S. entities towards international causes doubled in real terms [@DesaiKharas:2018, 505]. In 2018, this amounted to \$22.88 billion in individual donations earmarked specifically for international affairs INGOs [@GivingUSA:2019]. This turn to private donors opens INGOs up to new dynamics in seeking out funds. Many organizations that had previously relied on governments and foundations had optimized their fundraising strategies for those donors. But individual donors may behave differently from governments and foundations, so INGOs need to adapt their fundraising strategies to a different set of potential donors.

Individual donors have different preferences and rely on signals of effectiveness that differ from the benchmarks that large donor agencies rely on. While INGOs have control over organizational-level factors such as financial transparency and accountability that motivate donations, the political environment in the host countries INGOs work in---including the global crackdown on civil society that is causing many organizations to seek out private funding in the first place---is a structural factor that INGOs can rarely influence on their own. How do organizational elements of private donor decision-making interact with broader structural limits on INGO fundraising? How do individual donors in the U.S. feel about donating to legally besieged NGOs abroad? How do organizational-level factors such as financial transparency and accountability compare with more structural-level factors such as host country civic environments when deciding to donate?

Existing research shows that organizational accountability [@Becker:2018; @Tremblay-BoirePrakash:2017] and financial transparency [@HarrisNeely:2021; @SaxtonKuoHo:2012; @Vaccaro:2006; @VaccaroMadsen:2006] play an important role in motivating individual donors to donate to NGOs. We extend this literature by evaluating how these organization-level characteristics influence private donor behavior in an era of shrinking civic space. In considering this question, we engage with research on the consequences of closing civic space on NGOs and donors [@DupuyPrakash:2018; @RightSpringmanWibbels:2022; @FransenDupuyHinfelaar:2021; @SundstromHenrySperling:2022; @Herrold:2020; @ChaudhryHeiss:2021]. We argue that both organization characteristics---such as an NGO's issue area, funding sources, and transparency and accountability practices---and external factors like an NGO's host-country political environment act as heuristics that influence foreign private donors' preferences.

We fielded a conjoint experiment to a nationally representative sample of 1,016 adults in the US, which allows us to directly compare the relative impact of host government repression of NGOs and organizational financial transparency and accountability on individual donor preferences. The result of this experiment shows that both financial transparency and accountability increase individuals' willingness to donate to NGOs. Conversely, learning about host government criticism and crackdown against NGOs decreases the likelihood of donation by itself; however, organizational features such as financial transparency and accountability can protect against this dampening effect. These effect sizes are substantial: our key treatment effects have a magnitude of around ten percentage points in the predicted probability of choosing to donate to an NGO.

@@

A large body of work in philanthropy explores the factors that shape private donor motivation, including altruism, reputational benefits, and alignment with personal values [@BilodeauSlivinski:1997; @Wiepking:2010; @BekkersWiepking:2011]. However, this research has overwhelmingly looked at individual donor behavior and motivation for giving to *domestic* organizations instead of international causes [@BekkersWiepking:2011; @Tremblay-BoirePrakash:2017]. The geographic scope of an organization's mission matters to donors. Giving to local organizations involves fewer recipients and more observable results, while the benefits of donating to organizations working abroad are less visible and more removed from donors [@CasaleBaumann:2015]. In a survey experiment, @Tremblay-BoirePrakash:2017 confirm this, finding that "donors are more likely to donate to a charity operating locally than to a charity providing identical service abroad" [-@Tremblay-BoirePrakash:2017, 644]. This difference in donor preferences for international NGO services has been understudied because conventionally, states and foundations have been among the main funders of INGOs.

The third wave of democratization, the collapse of the Soviet Union, and the spread of a "pro-NGO norm" have led to an increased role and visibility for NGOs in recent decades [@Reimann:2006]. Donor states have channeled substantial resources through humanitarian and development NGOs, often in response to donor fears that recipient states would use aid inefficiently [@Dietrich:2013]. Institutional donors and private funders perceived NGOs as more efficient, more nimble, less bureaucratic, and more trustworthy than states that face poor governance and weak political institutions [@Chaudhry:2022], and as a result, NGOs have become crucial actors in both development and democratization [@Herrold:2020, 142--43].

But as INGOs became increasingly powerful actors in global policy circles, they faced a number of criticisms. Many organizations struggle with the conflicting imperatives of being large organizations that can operate efficiently at scale, while maintaining the grassroots contacts and consensus building that contributed to their early successes [@MitchellSchmitzBruno-vanVijfeijken:2020; @Jalali:2008; @Bano:2012; @Rahman:2006]. Many nonprofit scandals involve some form of financial wrongdoing [@PrakashGugerty:2010b]. Globally, around 5% of revenue is lost annually due to fraud [@ACFE:2021]. Misappropriation of assets and embezzlement schemes are among the most common forms of financial fraud [@LamotheTer-MkrtchyanRuddle:2023]. These criticisms and concerns, in turn, have led donors to demand greater NGO responsiveness, especially regarding financial transparency and accountability.

Assessing an NGO's deservingness and efficiency demands substantial time and resources, which individual donors typically lack [@CrosonShang:2011; @Tremblay-BoirePrakash:2017]. As such, these donors rely on various cues and heuristics when deciding to support an NGO. Nonprofits fundraising from individuals work to shape donor perceptions through the use of framing [@ChaudhryHeiss:2021]. We explore two possible categories of heuristic frames: (1) organization-level heuristics, reflected in NGO practices, and (2) structural heuristics, particularly an NGO's relationship with its host country. We hypothesize that each type plays a role in shaping individual preferences for engaging in philanthropy.

@@

One powerful heuristic for potential donors is organizational efficacy or the belief that an individual's donation will make a real difference [@ChaudhryHeiss:2021]. This can manifest itself through multiple mechanisms such as the amount an organization spends on overhead costs, whether it makes its finances public, or whether it participates in voluntary accountability practices, such as listing organizational details on websites like GuideStar and Charity Navigator.

Financial transparency, or the "degree of completeness of information provided by [organizations] to the [public] concerning [their] activities," [@VaccaroMadsen:2006, p. 146] is one tangible heuristic that donors turn to when making decisions decisions. NGOs can engage in financial transparency in a variety of ways such as distribution of audited financial statements and third party intermediaries. Research shows that transparency is associated with greater donations, lower debt, and better governance [@HarrisNeely:2021; @HarrisPetrovitsYetman:2023; @SaxtonKuoHo:2012]. It can also enhance donor trust and confidence in organizations [@Vaccaro:2006; @VaccaroMadsen:2006].

@@

@@

Accountability is another factor that can impact donor willingness to donate, and it is necessary for promoting public trust in NGOs. While transparency is an organization's willingness to make critical data about the organization public, accountability is the willingness of the organization to explain its action to shareholders [@CharityNavigator:2020]. It also encompasses how well an organization's mission is aligned with its resources and how it manages resources to optimize performance [@Sloan:2009]. As NGOs gain more prominence, donors demand more information about their activities and strategies [@Tremblay-BoirePrakash:2015]. Explaining actions to stakeholders can be important because many NGOs---even well-intentioned ones---can devote excessive resources towards marketing and overheads [@Bowman:2006]. Private donors, subsequently, tend to avoid NGOs that spend a large share of their revenues for overhead, non-mission-related purposes like administration, management and fundraising. The absence of accountability is associated with decreased public trust, worse reputation, and lower perceived quality [@Becker:2018].

Though individual donors would like to have some assurance that their resources will be used appropriately and organizations are not spending too much on overhead, individuals interested in supporting charities cannot thoroughly vet every aspect of the organization [@Tremblay-BoirePrakash:2017; @ChaudhryHeiss:2021]. To more easily judge organizational accountability, donors may use ratings by watchdog organizations like Charity Navigator and GuideStar as heuristics to guide their decisions. Research shows that initiatives to address accountability gaps by these organizations can substantially increase donations [@Sloan:2009; @GordonKnockNeely:2009; @Tremblay-BoirePrakash:2017].

> **H~1b~**: If NGOs are accountable and hold regular third-party audits, then individual private donors will have a higher likelihood of supporting or donating to them.

@@

While NGOs have control over organizational factors such as transparency and accountability and can work to improve these characteristics and public perception of them, NGOs may have little to no control over structural factors that can also influence philanthropy. These structural factors include political contexts in NGO host countries, especially their environments for civil society organizations. A global cascade of anti-NGO laws in recent decades has created barriers to entry, funding, and advocacy for civic organizations [@GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020; @Chaudhry:2022; @Heiss:2017; @DupuyRonPrakash:2016]. As a result, NGOs must spend more time, effort, and resources to ensure their survival---at the expense of pursuing their missions.

Some NGOs have adapted to repressive environments by recreating their organizations and changing their mission to avoid directly confronting the government. Analyzing the impact of anti-NGO laws in Bangladesh and Zambia, scholars have found that when faced with government restrictions, NGOs with broad missions restricted their strategies to service work and abandoned their advocacy work. Those that continued to engage in transnational advocacy changed the targets of their advocacy, the issues they focused on, and the language they used [@FransenDupuyHinfelaar:2021]. Similarly, in Russia, many NGOs working with foreign partners and lobbying the central government no longer consider it an effective strategy [@SundstromHenrySperling:2022]. Overall, fewer groups may end up focusing on contentious causes, instead focusing on regime-compatible causes.

The changes in civic environments impacting NGO activity have also influenced official donor responses. Scholars have found that while multilateral donors do not reduce aid in response to anti-NGO laws, these laws are associated with a 32% decline in bilateral aid inflows [@DupuyPrakash:2018]. Further, advocacy-oriented donors---i.e., those that fund democracy and civil society promotion activities---reduced their spending by 74%, but did not cut their spending on development projects such as education, health, water, and sanitation [@RightSpringmanWibbels:2022]. However, previous literature has established that individual private donors do not make the same considerations as official donors when deciding to donate [@DesaiKharas:2018; @ChaudhryHeiss:2021]. Private foreign aid does not have to deal with the same strategic and political considerations as official aid and therefore, may be in a better position to respond to recipients' needs on the grounds [@EasterlyWilliamson:2011]. Anti-NGO repression and civil society laws may act as a heuristic to individuals donors that NGOs undertake crucial work abroad, which is why governments perceive them as threatening and crack down on them. Therefore, restrictions imposed by host-country governments are likely to boost donors' inclination to donate to affected organizations.

@@

Identifying the causal link between structural factors and donor behavior is more difficult than measuring the link between organizational factors and donor behavior because (1) NGOs have less direct control over their political environments, and (2) organizational characteristics like financial transparency and accountability can sour the NGO-government relationship---governments may be more likely to target NGOs that lack financial transparency and accountability. Moreover, states pass anti-civil society laws in response to broader political trends within their borders, and they regulate NGOs strategically to maximize the benefits and minimize the costs of working with international NGOs [@Heiss:2017]. States are more likely to restrict NGOs when organizations pose a threat to regime stability or preferences—for instance, when INGO issue areas threaten government policies, or when INGOs receive substantial foreign funding [@DupuyRonPrakash:2015].

Donors show increased willingness to donate to NGOs facing government crackdown/criticism while being financial transparent and accountable. While concerns about NGO mismanagement and lack of financial transparency may signal concerns about an NGO's operations and its ability to abide by the host country's regulatory environment, the presence of such transparency and accountability may instead convince donors that government targeting of NGOs may be ill-intentioned rather than simply a manifestation of financial oversight. Donors may view anti-NGO laws and inhospitable civic environments as a sign of governments looking to crack down on groups that seek to keep governments accountable, rather than merely auditing and punishing groups that violate routine regulations.

Therefore, both states and individual donors expand and contract their regulatory environments (for states) and support (for donors) for NGOs in response to the interplay between domestic politics and organizational characteristics. The same NGO feature that can increase the likelihood of government crackdown can also simultaneously serve as a donor heuristic and influence perceptions of NGO deservingness.

@@

## Effect of crackdown

>>>>

---
title: "Why Donors Donate: Disentangling Organizational and Structural Heuristics for International Philanthropy"
short-title: "Why Donors Donate"
date: "2025-01-17"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  url: http://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
  orcid: 0000-0003-3506-7590
- name: Marc Dotson
  affiliation: Utah State University
  url: https://occasionaldivergences.com/
  email: marc.dotson@usu.edu
  orcid: 0000-0001-7846-3660
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  Governments across the world have increasingly used laws to restrict the work of nonprofits, which has led to a reduction in public or official foreign aid directed towards these groups. Many international nonprofits, in response, have turned to individual donors to offset the loss of traditional fundings. What are individual donors' preferences regarding donating to legally besieged nonprofits abroad? We conducted a conjoint experiment on a nationally representative sample of likely donors in the US and found that learning about host government criticism and restrictions on nonprofits decreases individuals' preferences to donate to them. However, organizational features such as financial transparency and accountability can protect against this dampening effect. Our results have important implications both for understanding private international philanthropy and how nonprofits can better frame their fundraising appeals at a time when they are facing restrictive civic spaces and hostile governments abroad.
additional-info: |
  **Data Availability:** All data and replication code are publicly available at <https://github.com/andrewheiss/silent-skywalk>
keywords:
  - philanthropy
  - conjoint experiments
  - donor heuristics
  - repression
  - NGOs
  - civil society
  - nonprofits
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/silent-skywalk>"
bibliography: references.bib
csl: csl/apa.csl
link-citations: true
word-count: "9,300"
---

@@

Non-governmental organizations (NGOs) are a core component of global governance and service delivery. Non-governmental organizations (NGOs) are a core component of global governance and service delivery. Over the last two decades, however, NGOs face restrictive civic spaces abroad [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020]. Repressive governments have increasingly used legal measures to create barriers to funding and advocacy for a variety of NGOs [@DupuyRonPrakash:2016; @Chaudhry:2022; @Heiss:2017]. For instance, in 2012 Russia passed its infamous "Foreign Agent" law (Federal Law No. 121-F2), which forced organizations receiving foreign funding and engaging in political activities to register as foreign agents, which forced hundreds of organizations to close down [@Kirova:2024]. These developments were not limited to authoritarian regimes. Hybrid and democratic states have also restricted NGOs' operations within their territories. Between 2014 and 2019, the Indian government banned 14,500 NGOs from receiving money from abroad, including prominent international groups such as the Ford Foundation and Greenpeace [@PressTrustofIndia:2019]. Policymakers have referred to this phenomenon as the "closing of civic space" or "shrinking civic space" [@CarothersBrechenmacher:2014; @Rutzen:2015b]. Data from the CIVICUS Civil Society Monitor shows that civil society groups face serious restrictions in more than 60% of countries---117 in total [@CIVICUSMonitor]. This phenomenon of closing civic space has a large deterrent effect on both advocacy *and* service-delivery NGOs [@SpringmanMaleskyRight:2022].

These restrictions have adversely affected the ability of both domestic and international NGOs to operate and raise funds. Official aid funds have decreased substantially in countries that have imposed restrictive anti-NGO laws [@ChaudhryHeiss:2018; @DupuyPrakash:2018] and official donors channel fewer funds to contentious issues such as human rights, elections, and advocacy, and instead choose to fund "tame" causes bush as health, education, and development, among others [@ChaudhryHeiss:2018; @Bush:2015; @Herrold:2020; @RightSpringmanWibbels:2022]. In response to this dramatic shift in the NGO funding landscape, international NGOs (INGOs) have increasingly turned to individual donors abroad to offset the loss of funding from government and foundation sources [@BanksHulmeEdwards:2015].

The amount of private foreign aid is not insignificant---from 2000--2010, private giving by US entities towards international causes doubled in real terms [@DesaiKharas:2018, 505]. In 2018, this amounted to \$22.88 billion in individual donations earmarked specifically for international affairs INGOs [@GivingUSA:2019]. This turn to private donors opens INGOs up to new dynamics in seeking out funds. Individual donor behavior is unlike official donor behavior [@DesaiKharas:2018]. They have different preferences---while official donors channel funds to countries with better institutional quality or to reward governmental performance, [@Bermeo:2011; @DietrichWright:2015], individual donors are more inclined to donate to NGOs in countries experiencing a humanitarian crisis [@DesaiKharas:2018]. As such, individual donors may have separate metrics of effectiveness or impact that can differ from the benchmarks that large donor agencies rely on, and INGOs may need to adapt their fundraising strategies to a different set of potential donors.

While INGOs have control over organizational-level factors that determine their effectiveness, the political environment in the host countries INGOs work in---including the global restrictions on civil society that is causing many organizations to seek out private funding in the first place---is a structural factor that INGOs can rarely influence on their own. How do individual donors in the US feel about donating to legally besieged NGOs abroad? How do organizational elements of private donor decision-making interact with broader structural limits on INGO fundraising? While there are many organizational-level factors that can determine an NGO's impact, including such as organizational structure, financial transparency, accountability, leadership, and strategy [@MitchellSchmitzBruno-vanVijfeijken:2020], in this paper, we focus on the two factors that are most visible to donors outside of the organizations and on which information is available publicly---financial transparency and accountability.

Existing research shows that organizational accountability [@Becker:2018; @Tremblay-BoirePrakash:2017]  and financial transparency [@HarrisNeely:2021; @SaxtonKuoHo:2012; @Vaccaro:2006; @VaccaroMadsen:2006] play an important role in motivating individual donors to donate to NGOs. We extend this literature by evaluating how these organization-level characteristics influence private donor behavior in an era of shrinking civic space. In considering this question, we engage with research on the consequences of closing civic space on NGOs and donors [@DupuyPrakash:2018; @RightSpringmanWibbels:2022; @FransenDupuyHinfelaar:2021; @SundstromHenrySperling:2022; @Herrold:2020; @ChaudhryHeiss:2021]. We argue that organization characteristics---such as an NGO's transparency and accountability practices---can serve as important heuristics that influence foreign private donors' preferences even during closing civic space.

We fielded a conjoint experiment to a nationally representative sample of 1,016 adults in the US. This experiment allows us to assess individual level giving preferences among likely donors in the US towards NGOs abroad working in a variety of issue areas---human rights, environment, emergency response, and refugee relief.[^not-elite] More importantly, it allows us to directly compare the relative impact of host government restriction of NGOs and organizational financial transparency and accountability on individual donor preferences. The result of this experiment shows that both financial transparency and accountability increase individuals' willingness to donate to NGOs. Conversely, learning about host government criticism of and restrictions on NGOs decreases the likelihood of donation by itself; however, organizational features such as financial transparency and accountability can protect against this dampening effect.

[^not-elite]: To be clear, we do not examine elite-level giving or philanthropy by wealthy donors in the US.

@@

A large body of work in philanthropy explores the factors that shape private donor motivation, including altruism, reputational benefits, and alignment with personal values [@BilodeauSlivinski:1997; @Wiepking:2010; @BekkersWiepking:2011]. However, this research has overwhelmingly looked at individual donor behavior and motivation for giving to *domestic* organizations instead of international causes [@BekkersWiepking:2011; @Tremblay-BoirePrakash:2017]. Giving to local organizations involves more observable results, while the benefits of donating to organizations working abroad are less visible and more removed from donors [@CasaleBaumann:2015]. @Tremblay-BoirePrakash:2017 confirm this, finding that "donors are more likely to donate to a charity operating locally than to a charity providing identical service abroad" [-@Tremblay-BoirePrakash:2017, 644].

This difference in donor preferences for giving to international NGOs has been understudied because conventionally, states and foundations have been among the main funders of INGOs. The third wave of democratization and the collapse of the Soviet Union led donor states to channel substantial resources through NGOs, often in response to donor fears that recipient states would use aid inefficiently [@Dietrich:2013]. Institutional donors and private foundations perceived NGOs as more efficient, more nimble, less bureaucratic, and more trustworthy than states that face poor governance and weak political institutions.

But as INGOs became increasingly powerful actors in global policy circles, they faced a number of criticisms. Many organizations struggle with the conflicting imperatives of being large organizations that can operate efficiently at scale, while maintaining the grassroots contacts and consensus building that contributed to their early successes [@MitchellSchmitzBruno-vanVijfeijken:2020; @Jalali:2008; @Bano:2012; @Rahman:2006]. In response, many nonprofits sought to establish and enhance their accountability, responsiveness, and legitimacy [@edwardshulme; @GourevitchLake:2012]. These actions were a result of many organizational-level factors impact nonprofit efficacy, including organizational structure, leadership, strategy, transparency and accountability. Organizational and governance structures define how decisions are made and who participates in decision making. While some nonprofits follow more democratic governance models---such as expanded membership to Global South countries, devolving more decision-making powers to the regional or country levels, others have a more centralized structure [@Wong:2012; @MitchellSchmitzBruno-vanVijfeijken:2020, 155]. Leadership can be defined as how leaders "handle the external and internal politics of making their organizations more effective and responsive," and what sort of organizational cultures they create [@MitchellSchmitzBruno-vanVijfeijken:2020, 178]. Finally, strategy is whether the organization has a long-term strategic plan that emphasizes sustainability, innovation, learning and adaptation [@MitchellSchmitzBruno-vanVijfeijken:2020, 100].

However, organizational structure, leadership, and strategy (among other organizational-level factors) are often not visible to the average individual (non-elite) donor, especially in the context of international NGOs. Therefore, in this paper, we focus on the two organizational-level factors that are most visible to the average donor---financial transparency and accountability. Assessing an NGO's deservingness and efficiency demands substantial time and resources, which individual donors typically lack [@CrosonShang:2011; @Tremblay-BoirePrakash:2017]. As such, these donors rely on various easily available cues and heuristics when deciding to support an NGO. News about financial mismanagement and lack of transparency and accountability may be more easily available to these donors. Many nonprofit scandals involve some form of financial wrongdoing [@PrakashGugerty:2010b]. Notable examples include the Oil-for-Food program, initiated by the United Nations in 1996 to allow Iraq to sell enough oil to pay for food and other necessities for its population, which has been suffering from sanctions imposed on Iraq since the First Gulf War. However, in 2004, the widespread financial fraud by the program, limited distribution of food by nonprofits and lack of accountability was widely publicized. Numerous other such cases---covered in @GibelmanGelman:2001 and @GibelmanGelman:2004---show how financial mismanagement, embezzlement, fraud, and lack of accountability can decrease public trust in and donations towards these nonprofits. Globally, around 5% of revenue is lost annually due to fraud [@ACFE:2021]. Misappropriation of assets and embezzlement schemes are among the most common forms of financial fraud [@LamotheTer-MkrtchyanRuddle:2023]. To save time and present easily accessible information, watch dogs such as Charity Navigator and Better Business Bureau Wise Giving Alliance often rate charities on financial and accountability metrics and present this information in a quick, user-friendly format.

Thus, in this paper, we explore the effects of two possible categories of heuristic frames that can impact individual-level giving: (1) organization-level heuristics, reflected in practices of transparency and accountability, and (2) structural heuristics, particularly an NGO's relationship with its host country. We hypothesize that each type plays a role in shaping individual preferences for engaging in philanthropy.

@@

We first examine the impact of two organizational practices---financial transparency and accountability---on individual donor preferences. Financial transparency, or the "degree of completeness of information provided by [organizations] to the [public] concerning [their] activities," [@VaccaroMadsen:2006, p. 146] is one tangible heuristic that donors turn to when making decisions. Transparency can involve proactive sharing of information with stakeholders, participation in data transparency platforms, avoiding financial misconduct, and exhibiting trustworthiness. NGOs can engage in financial transparency in a variety of ways such as distribution of audited financial statements and voluntarily sharing information with third party intermediaries. Research shows that transparency is associated with greater donations, lower debt, and better governance [@HarrisNeely:2021; @HarrisPetrovitsYetman:2023; @SaxtonKuoHo:2012]. It can also enhance donor trust and confidence in organizations [@Vaccaro:2006; @VaccaroMadsen:2006].

@@

Accountability is another factor that can impact donor preferences. Accountability is the willingness of the organization to explain its action to shareholders [@CharityNavigator:2020] and build and manage relationships with stakeholders or client populations and donor groups. It also encompasses how well an organization's mission is aligned with its resources [@Sloan:2009]. As NGOs gain more prominence, stakeholders, or "any group or individual who can affect or is affected by the achievement of an organization's objectives," [@MitchellAgleWood:1997, 854] want to be assured that their donations or support or support of an organization will make a real difference. More effective accountability can encourage NGOs to become more closely aligned with community practices, provide greater assurances to stakeholders and help expand their support [@Murtaza:2012]. Problems with accountability usually occur when organizations ignore their stakeholders' preferences and use judgements or actions which may not align with one or more of their constituent groups [@McCambridge:2019]. The absence of accountability is associated with decreased public trust, worse reputation, and lower perceived quality [@Becker:2018].

Though individual donors would like to have some assurance that their resources will be used appropriately and organizations are not spending too much on overhead, individuals interested in supporting charities cannot thoroughly vet every aspect of the organization [@Tremblay-BoirePrakash:2017; @ChaudhryHeiss:2021]. To more easily judge financial transparency and accountability, donors may use ratings by watchdog organizations like Charity Navigator and GuideStar as heuristics to guide their decisions.[^cn-multiple] Research shows that initiatives to address information gaps by these organizations can substantially increase donations [@Sloan:2009; @GordonKnockNeely:2009; @Tremblay-BoirePrakash:2017].

@@

> **H~1b~**: If NGOs are accountable, then individual private donors will have a higher likelihood of supporting or donating to them.

[^cn-multiple]: For instance, Charity Navigator lists multiple different accountability and financial metrics in an easily available user-friendly format on its website: <https://www.charitynavigator.org/about-us/our-methodology/ratings/accountability-finance/>

@@

While NGOs have control over organizational factors such as transparency and accountability and can work to improve these characteristics and public perception of them, NGOs may have little to no control over structural factors that can also influence philanthropy. These structural factors include political contexts in NGO host countries, especially a restrictive environment for civil society organizations. A global cascade of anti-NGO laws in recent decades has created barriers to entry, funding, and advocacy for civic organizations [@GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020; @Chaudhry:2022; @Heiss:2017; @DupuyRonPrakash:2016]. The use of such anti-NGO laws, also referred to as administrative crackdown [@Chaudhry:2022], is different from regulations that simply set standards for appropriate organizational behavior and set penalties for violations [@North:1990].[^violent-crackdown] However, the anti-NGO laws discussed in this paper, such as Russian Foreign Agent law, are not grounded in international legal principles, and do not give nonprofits the freedom of association or the freedom to freely access human and material resources [@Voule:2023; @Robinson:2024]. As a result, NGOs must spend more time, effort, and resources to ensure their survival---at the expense of pursuing their missions.

[^violent-crackdown]: Crackdowns can also be violent. However, unlike administrative crackdowns, violent crackdowns are often hard to attribute to governments, since governments often outsource repression. Moreover, given the ubiquity of anti-NGO laws in recent decades, we focus on the impact of these legal crackdowns on donor preferences.

Some NGOs have adapted to these legal crackdowns by recreating their organizations and changing their mission to avoid directly confronting the government. Analyzing the impact of anti-NGO laws in Bangladesh and Zambia, scholars have found that when faced with government restrictions, NGOs with broad missions restricted their strategies to service work and abandoned their advocacy work. Those that continued to engage in transnational advocacy changed the targets of their advocacy, the issues they focused on, and the language they used [@FransenDupuyHinfelaar:2021]. Similarly, in Russia, many NGOs working with foreign partners and lobbying the central government no longer consider it an effective strategy [@SundstromHenrySperling:2022]. Overall, fewer groups may end up focusing on contentious causes, instead focusing on issues that are less likely to invoke the ire of the host government, including in areas such as health and education.

The changes in civic environments impacting NGO activity have already influenced official donor responses. Scholars have found that while multilateral donors do not reduce aid in response to anti-NGO laws, these laws are associated with a 32% decline in bilateral aid inflows [@DupuyPrakash:2018]. Further, advocacy-oriented donors---i.e., those that fund democracy and civil society promotion activities---reduced their spending by 74%, but did not cut their spending on development projects such as education, health, water, and sanitation [@RightSpringmanWibbels:2022].

However, previous literature has established that individual private donors do not make the same considerations as official donors when deciding to donate [@DesaiKharas:2018; @ChaudhryHeiss:2021]. Private foreign aid does not have to deal with the same strategic and political considerations as official aid and therefore, may be in a better position to respond to recipients' needs on the grounds [@EasterlyWilliamson:2011]. When a potential recipient NGO faces legal crackdown in its host country, private individual donors can respond by (1) increasing their donations as a sign of support and solidarity, (2) decreasing their donations, punishing the NGO for doing something to run afoul of its host government, or (3) not considering the host country legal environment and making no change in their donation behavior. @ChaudhryDotsonHeiss:2021 explore how different combinations of donor characteristics determine which response individuals are likely to take when NGOs face crackdown. They find that donors with longer experiences with the nonprofit sector and high levels of social trust---i.e. those who frequently volunteer, regularly donate to charity, and trust political institutions---are more likely to maintain their support for international NGOs that face criticism or crackdown abroad. Thus, anti-NGO laws may act as a heuristic to individuals donors that NGOs undertake crucial work abroad, which is why governments perceive them as threatening and seek to restrict their work.

This is especially true for nonprofits that rely more heavily on private funds, rather than government funds. Previous research finds that, "individual donors thus seem to be more willing to support besieged human rights organizations when they are unencumbered by government funds." [@ChaudhryHeiss:2021, 496]. Similar trends can be seen even after nonprofits face restrictive political and civic environments in other specific issue areas. For instance, in 2022, after the Supreme Court overturned Roe v. Wade, private philanthropy towards abortion funds skyrocketed [@PBSNews:2022]---a significant development as abortion funds rely largely on individual donors [@Cohen:2022]. Therefore, we expect that anti-NGO restrictions imposed by host-country governments are likely to boost individual donors' inclination to donate to affected organizations.

@@

Identifying the causal link between structural factors and donor behavior is more difficult than measuring the link between organizational factors and donor behavior because (1) NGOs have less direct control over their political environments, and (2) organizational characteristics like financial transparency and accountability can sour the NGO-government relationship---governments may be more likely to target NGOs that lack financial transparency and accountability. Moreover, states pass anti-civil society laws in response to broader political trends within their borders, and they regulate NGOs strategically to maximize the benefits and minimize the costs of working with international NGOs [@Heiss:2017]. States are more likely to restrict NGOs when organizations pose a threat to regime stability or preferences---for instance, when INGO issue areas threaten government policies, or when INGOs receive substantial foreign funding [@DupuyRonPrakash:2015].

Donors show increased willingness to donate to NGOs facing government restrictions/criticism while being financial transparent and accountable. While concerns about NGO mismanagement and lack of financial transparency may signal concerns about an NGO's operations and its ability to abide by the host country's regulatory environment, the presence of such transparency and accountability may instead convince donors that government targeting of NGOs may be ill-intentioned rather than simply a manifestation of financial oversight. Donors may view anti-NGO laws and inhospitable civic environments as a sign of governments looking to crack down on groups that seek to keep governments accountable, rather than merely auditing and punishing groups that violate routine regulations.

Therefore, both states and individual donors expand and contract their regulatory environments (for states) and support (for donors) for NGOs in response to the interplay between domestic politics and organizational characteristics. The same NGO feature that can increase the likelihood of government restriction can also simultaneously serve as a donor heuristic and influence perceptions of NGO deservingness.

@@

## Effect of restrictions
